Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-23 paper.tex

For $\gamma>1$ the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \Omega_k^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisified in a
thin disk, the disk would be stable.

>>>>

For $\gamma>1$ the disk is stably stratified with a real vertical
buoyancy frequency $N_z$.  In a thin disk,     
\begin{align}\label{nz_thin}
  N_z^2 \simeq \left(\frac{\gamma-1}{\gamma}\right) \left(\frac{z}{H}\right)^2
  \Omega_k^2.  
\end{align}
This provides a stabilizing effect as buoyancy opposes vertical
motion, which is required for the VSI. In fact, if we consider
axisymmetric adiabatic disturbances, then because the 
Solberg-Hoiland criteria \citep{tassoul78} are usually satisfied in a
thin disk, the disk would be stable.

### 2015-04-23 setup.tex

\section{Governing equations and disk models}\label{setup}
We consider  an inviscid, non-self-gravitating three-dimensional (3D)
disk orbiting a central star of mass $M_*$. We adopt cylindrical
co-ordinates $(r,\phi, z)$ centered on the star to describe the global
disk. The governing fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot{\left(\rho\bm{v}\right)} = 0,\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi_*,\\
  &\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  = - \gamma P
  \nabla\cdot\bm{v} - \Lambda % \frac{\rho}{t_c}\left(\frac{P}{\rho} -
    % \left.\frac{P}{\rho}\right|_{t=0}\right)
  , \label{full_energy}
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the velocity field (the
rotation frequency being $\Omega=v_\phi/r$), $P$
is the pressure, $\gamma$ is the constant adiabatic index, and $\Phi_*
= -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational potential of the
central star and $G$ is the gravitational constant. 
In the energy equation (Eq. \ref{full_energy}) the source term
$\Lambda$ represent non-adiabatic effects which we
model below. The gas temperature $T$ is defined through the ideal gas 
equation of state,  
\begin{align}
P = \frac{\mathcal{R}}{\mu}\rho T,
\end{align}
where $\mathcal{R}$ is the gas constant and $\mu$ is the mean molecular
weight.

\subsection{Thermal relaxation}
We model non-adiabatic effects by defining $\Lambda$ as a thermal
relaxation term,
\begin{align}\label{thermal_relax}
  \Lambda  \equiv \frac{\rho\mathcal{R}}{\mu t_c}\left(T - T_0\right),
\end{align}
where $T_0=T_0(r,z)$ is the initial temperature, and $t_c$ is a
thermal relaxation timescale parametrized via $\beta$ such that 
\begin{align}\label{beta_def}
  t_c = \beta \Omega_k^{-1},
\end{align}
where $\Omega_k=\sqrt{GM_*/r^3}$ is the Keplerian frequency.

For most of this paper we will take $\beta$ as a constant input
parameter, so that we can vary the thermodynamic response of the
disk from isothermal ($\beta\ll 1$) to adiabatic ($\beta \gg 1$) in a
controlled manner. However, for application to PPDs in 
\S\ref{application} we will model $\beta$ based on physical thermal
timescales, in which case $\beta$ is non-constant and is determined by
the equilibrium disk properties and the perturbation lengthscales of
interest (e.g. Fig. \ref{bcrit_mmsn2d}).

>>>>

\section{Governing equations and disk models}\label{setup}
We consider  an inviscid, non-self-gravitating three-dimensional (3D)
disk orbiting a central star of mass $M_*$.  Cylindrical
co-ordinates $(r,\phi, z)$ are centered on the star. The governing fluid equations are
\begin{align}
  &\frac{\p\rho}{\p t} + \nabla\cdot{\left(\rho\bm{v}\right)} = 0,\\
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} =
  -\frac{1}{\rho}\nabla P - \nabla\Phi_*,\\
  &\frac{\p P}{\p t} + \bm{v}\cdot\nabla P  = - \gamma P
  \nabla\cdot\bm{v} - \Lambda % \frac{\rho}{t_c}\left(\frac{P}{\rho} -
    % \left.\frac{P}{\rho}\right|_{t=0}\right)
  , \label{full_energy}
\end{align}
where $\rho$ is the mass density, $\bm{v}$ is the velocity field (the
rotation frequency being $\Omega=v_\phi/r$), $P$
is the pressure, $\gamma$ is the constant adiabatic index, and $\Phi_*
= -GM_*(r^2 + z^2)^{-1/2}$ is the gravitational potential of the
central star and $G$ is the gravitational constant. 
In the energy equation (Eq.\ \ref{full_energy}) the sink term
$\Lambda$ includes non-adiabatic cooling and (negative) heating. 
The gas temperature $T$ obeys the ideal gas 
equation of state,  
\begin{align}
P = \frac{\mathcal{R}}{\mu}\rho T,
\end{align}
where $\mathcal{R}$ is the gas constant and $\mu$ is the mean molecular
weight.

\subsection{Thermal relaxation}
We model radiative effects as thermal
relaxation with a timescale $t_c$, i.e.,
\begin{align}\label{thermal_relax}
  \Lambda  \equiv \frac{\rho\mathcal{R}}{\mu}\frac{T - T_0}{t_c},
\end{align}
where $T_0=T_0(r,z)$ is the equilibrium temperature.  We 
define the dimensionless cooling time 
\begin{align}\label{beta_def}
  \beta \equiv \Omega_k t_c
\end{align}
relative to the Keplerian frequency, $\Omega_k=\sqrt{GM_*/r^3}$.

For most of this paper we will take $\beta$ as a constant input
parameter, so that we can vary the thermodynamic response of the
disk from isothermal ($\beta\ll 1$) to adiabatic ($\beta \gg 1$) in a
controlled manner.

In \S\ref{application} we analyze the VSI using a realistic model for 
$\beta$ in PPDs with a dust opacity, using Eq.\ (\ref{beta_mmsn_simp}).  
 Fig. \ref{bcrit_mmsn2d} shows that in this model $\beta >> 1$ in the 
 dense inner disk and near the midplane, but that rapid cooling ($\beta \ll 1$)
 is possible elsewhere.  Furthermore short wavelength perturbations can cool 
 quicker ...

 \emph{TODO: $k_x$ is in Fig. \ref{bcrit_mmsn2d} but not explained yet...}
% based on physical thermal
%timescales, in which case $\beta$ is non-constant and is determined by
%the equilibrium disk properties and the perturbation lengthscales of
%interest (e.g. Fig. \ref{bcrit_mmsn2d}).
